Appendix S1. Linear regression models and non-linear population dynamics

### 2024-05-02 Appendix_S1_v01.Rmd
Env-dependent growth

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE}
library(Hmsc)
library(ggplot2)
library(tidyr)
library(coda)
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
```

@@

$$ N_{t+1} = \beta_0 + \beta_1 N_t + \epsilon_t $$
```{r,warning=FALSE}
# Fit the model
m.1.ar <- arima(x = log(N), order = c(1,0,0), include.mean = T, method="CSS")
m.1.lm <- lm(log(dat$N[2:t]) ~ log(dat$N[1:(t-1)]))
#plotting the series along with the fitted values
m.1.ar.fit <- log(N) - residuals(m.1.ar)
m.1.lm.fit <- log(dat$N[2:t]) - m.1.lm$resid
dat$ar1.fit <- m.1.ar.fit
dat$lm.fit <- NA
dat$lm.fit[2:t] <- m.1.lm.fit
```
```{r figs2, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line).",warning=FALSE}
colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
ggplot(data=dat,aes(x=time)) +
  geom_point(aes(y=log(N),color="data")) +
  geom_point(aes(y=ar1.fit, color = "AR1 fit"), alpha=0.5) +
  geom_point(aes(y=lm.fit, color = "lm fit"), alpha=0.5) +
  geom_line(aes(y=log(N), color="data"), linewidth=.4) +
  geom_line(aes(y=ar1.fit, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
  geom_line(aes(y=ar1.fit, color = "lm fit"), alpha=0.5, linewidth=.4) +
  theme_bw() +
  xlab("time") +
  ylab("N") +
  labs(color="Legend") +
  scale_color_manual(values = colors)
```

@@

```{r figs5, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
Gradient <- constructGradient(m.1.sample,focalVariable="x",ngrid=29)
predY <- predict(m.1.sample,Gradient=Gradient,expected=TRUE)
#preds <-computePredictedValues(m.1.sample)
plotGradient(m.1.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="N_t",ylab="predicted N_t+1")
```

@@

Here, $\hat{W}$ is the maximal population growth rate (set to )

>>>>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE}
library(Hmsc)
library(ggplot2)
library(tidyr)
library(coda)
library(patchwork)
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
```

@@

$$ N_{t+1} = \beta_0 + \beta_1 N_t + \epsilon_t $$
```{r,warning=FALSE}
# Fit the model
m.1.ar <- arima(x = log(N), order = c(1,0,0), include.mean = T, method="CSS")
m.1.lm <- lm(log(dat$N[2:t]) ~ log(dat$N[1:(t-1)]))
#plotting the series along with the fitted values
m.1.ar.fit <- log(N) - residuals(m.1.ar)
m.1.lm.fit <- log(dat$N[2:t]) - m.1.lm$resid
dat$ar1.fit <- m.1.ar.fit
dat$lm.fit <- NA
dat$lm.fit[2:t] <- m.1.lm.fit
```
```{r figs2, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line).",warning=FALSE}
colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
ggplot(data=dat,aes(x=time)) +
  geom_point(aes(y=log(N),color="data")) +
  geom_point(aes(y=ar1.fit, color = "AR1 fit"), alpha=0.5) +
  geom_point(aes(y=lm.fit, color = "lm fit"), alpha=0.5) +
  geom_line(aes(y=log(N), color="data"), linewidth=.4) +
  geom_line(aes(y=ar1.fit, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
  geom_line(aes(y=lm.fit, color = "lm fit"), alpha=0.5, linewidth=.4) +
  theme_bw() +
  xlab("time") +
  ylab("N") +
  labs(color="Legend") +
  scale_color_manual(values = colors)
```

@@

```{r figs5, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
Gradient <- constructGradient(m.1.sample,focalVariable="x",ngrid=29)
predY <- predict(m.1.sample,Gradient=Gradient,expected=TRUE)
#preds <-computePredictedValues(m.1.sample)
plotGradient(m.1.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="N_t",ylab="predicted N_t+1")
```

@@

Here, $\hat{W}$ is the maximal population growth rate (set to `r r1.0` as above), $E$ is the local environmental trait optimum value, and $x_{i,t}$ is species *i* trait value at time *t*. We see that if $E = x_{i,t}$ then the growth rate is at the value $r$ = `r r1.0`. Here, we begin with $E = x_{i,t} = 0.8$, then simulate the environment $E$ value fluctuating randomly over time, and finally use a linear model to fit $E$ as a covariate.

```{r}
# Simulate initial species population growth with environment fluctuations
N1.0 <- 10
r1.0 <- 1.67
alpha.11 <- 0.00125
E.0 <- 0.8
x1.0 <- 0.8
# model function
disc_log_E <- function(r, N0, alpha, E, x) {
    Nt1 <- ((r*exp(-(E-x)^2))*N0) / (1+alpha*N0)
    return(Nt1)
}
# Simulation of model for t time steps
t <- 40
N <- rep(NA, t)
N[1] <- N1.0
E <- rep(NA, t)
E[1] <- E.0
for (i in 2:t) {
  N[i] <- disc_log_E(r=r1.0, N0=N[i-1], alpha=alpha.11, E=E[i-1], x=x1.0)
  E[i] <- E[i-1] + rnorm(1,0,.1)
}
```
```{r figs6, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs1}Population size \\textit{N} over time \\textit{t} for a discrete-time logistic growth model, with parameters $r_i$ = 1.67, $N_{1,0}$ = 14, and $\\alpha_{11}$ = 0.00125."}
# Plot simulation: ggplot
dat <- as.data.frame(cbind(N,E))
dat$time <- 1:t
p1 <- ggplot2::ggplot(dat, aes(time, N)) + geom_point() + geom_hline(yintercept = ((r1.0 - 1)/alpha.11),linetype = "dashed", color = "gray")
p2 <- ggplot2::ggplot(dat, aes(time, E)) + geom_point()
p1 + p2
```

## Linear statistical model with environmental covariate

We now include environment $E$ as a covariate in the linear model:

$$ N_{t} = \beta_0 + \beta_1 N_{t-1} + \beta_2 E_{t-1}+ \epsilon_t $$

```{r,warning=FALSE}
# Fit the model
m.2.ar <- arima(x = log(N), order = c(1,0,0), include.mean = T, method="CSS", xreg=E)
m.2.lm <- lm(log(dat$N[2:t]) ~ log(dat$N[1:(t-1)]) + E[1:(t-1)])
#plotting the series along with the fitted values
m.2.ar.fit <- log(N) - residuals(m.2.ar)
m.2.lm.fit <- log(dat$N[2:t]) - m.2.lm$resid
dat$ar2.fit <- m.2.ar.fit
dat$lm2.fit <- NA
dat$lm2.fit[2:t] <- m.2.lm.fit
```
```{r figs7, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs2}Population size over time (black line) with fitted values from a first-order autoregressive model (red dashed line).",warning=FALSE}
colors <- c("data" = "black", "AR1 fit" = "gray", "lm fit"="orange")
ggplot(data=dat,aes(x=time)) +
  geom_point(aes(y=log(N),color="data")) +
  geom_point(aes(y=ar2.fit, color = "AR1 fit"), alpha=0.5) +
  geom_point(aes(y=lm2.fit, color = "lm fit"), alpha=0.5) +
  geom_line(aes(y=log(N), color="data"), linewidth=.4) +
  geom_line(aes(y=ar2.fit, color = "AR1 fit"), alpha=0.5, linewidth=.4) +
  geom_line(aes(y=lm2.fit, color = "lm fit"), alpha=0.5, linewidth=.4) +
  theme_bw() +
  xlab("time") +
  ylab("N") +
  labs(color="Legend") +
  scale_color_manual(values = colors)
```
The linear model is a good fit when including the environmental covariate. $N_{t+1}$ and $N_t$ are no longer well-represented by a linear function without taking $E$ into account.

```{r figs8, fig.pos = "H",echo=FALSE, fig.show="hold", fig.cap = "\\label{fig:figs3}Population size (logarithm) at one time step $N_{t+1}$ as a function of log-population size in the previous time step $N_t$. "}
dN <- rep(NA, (t-1))
for (i in 1:(t-1)) {
  dN[i] <- log(N[i+1])-log(N[i])
}
d.dN <- as.data.frame(dN)
d.dN$Nt <- log(dat$N[1:(t-1)])
d.dN$Nt_plus_1 <- log(dat$N[2:t])
p1 <- ggplot2::ggplot(d.dN, aes(x=Nt,y=Nt_plus_1)) + geom_point()

d.dN <- as.data.frame(dN)
d.dN$Nt <- log(dat$N[1:(t-1)])
d.dN$Nt_plus_1 <- log(dat$N[2:t])
d.dN$E <- dat$E[2:t]
p2 <- ggplot2::ggplot(d.dN, aes(x=E,y=Nt_plus_1)) + geom_point()
p3 <- ggplot2::ggplot(d.dN, aes(Nt, dN)) + geom_point()
p4 <- ggplot2::ggplot(d.dN, aes(E, dN)) + geom_point()

p1+p2+p3+p4
```

## Bayesian linear statistical model: HMSC
We can estimate the same model parameters using HMSC:

```{r}
# prepare data in HMSC format
Y <- as.matrix(log(dat$N[2:t]))
XData <- data.frame(x=cbind(log(dat$N[1:(t-1)])),E[1:(t-1)])
colnames(XData)[2] <- "E"
m.2.hmsc <- Hmsc(Y=Y,XData=XData,XFormula=~x+E)
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 1000
transient <- 500*thin
verbose <- 500*thin
# sample MCMC
m.2.sample <- sampleMcmc(m.2.hmsc,thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
```

```{r}
m2.post.hmsc <- convertToCodaObject(m.2.sample)
summary(m2.post.hmsc$Beta)
plot(m2.post.hmsc$Beta)
```

```{r}
# AR1 coefficients (recall that the intercept is the term below multipled by 1 - phi1)
m.2.ar$coef
m.2.ar$coef[2]*(1-m.2.ar$coef[1])
# linear model
summary(m.2.lm)$coefficients[1:3,1:2]
# Bayesian estimates
summary(m2.post.hmsc$Beta)$statistics[1:3,1:2]
```

We recall that the interpretation of the coefficients in an arimaX (arima with covariates) model is difficult. They do not give the impact on $N_t$ per unit increase in X as in a regression. So we do not interpret the causation implied by the coefficient in the arimaX model. In the regression model, we can see that $E$ has a positive impact on $N_t$.

```{r figs9, fig.pos = "H", fig.show="hold", fig.cap = "\\label{fig:figs5} Observed (grey) and model-fit (blue) values for population size at time t (x-axis) and t+1 (y-axis).",results='hide',fig.keep='all'}
Gradient <- constructGradient(m.2.sample,focalVariable="E",non.focalVariables=list(x=list(2)),ngrid=39)
predY <- predict(m.2.sample,Gradient=Gradient,expected=TRUE)
#preds <-computePredictedValues(m.1.sample)
plotGradient(m.2.sample,Gradient,pred=predY,showData=T,measure="Y",main="",xlab="E_t",ylab="predicted N_t+1")
```

## Conclusions
In this example, the linear regression again works well to describe the impact of $E_t$ for $N_t$. We can see the response of course depends on the population size value, as the gradient plot is generally linear

a first-order auto-regressive model works well, bypassing the need to estimate logistic growth parameters $r_i$ and $\alpha_{ii}$. The density-dependence dynamics ($\Delta N \sim f(N_t)$) show an overall declining trend over time. The Bayesian estimation implemented in HMSC gives good parameter estimates.

```{r}
blah <- data.frame(cbind(log(dat$N[2:t]),log(dat$N[1:(t-1)]),E[1:(t-1)]))
colnames(blah) <- c("Nt1","Nt","E")
mod <- lm(Nt1 ~ Nt + E,data=blah)
jtools::effect_plot(mod,pred=E,plot.points = TRUE)
```

# Simulate initial species population growth
N1.0 <- rpois(1,10)
r1.0 <- 1.67
alpha.11 <- 0.00125
